---
title: "Betacoronavirus link prediction via Elmasri et al. approach"
author: "Maxwell J. Farrell"
date: "`r format(Sys.time(), '%B %d %Y')`"
output: 
  html_document:
    highlight: default
    toc: yes
    toc_depth: 2
    toc_float: true
    theme: yeti
urlcolor: blue
---

```{r setup, include = FALSE, eval=TRUE, echo=TRUE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

```{r loadinglibraries}
# devtools::install_github("melmasri/HPprediction")
require(HPprediction)
require(ape)
require(dplyr)
require(parallel)
```

```{r custom_funcs, eval=T, echo=F}
# Custom functions to parse predictions

# Modified topPairs function that attaches names from Z matrix
topPairs<-function(P,Z,topX=20){
    ## Plotting the top uncodumented interactions
    P[Z==1]<--1
    rownames(P) <- rownames(Z)
    colnames(P) <- colnames(Z)
    aux = melt(P)
    aux = aux[order(aux$value,decreasing=TRUE),]
    colnames(aux)<-c('Host', 'Parasite', 'p(interaction)')
    aux
    aux[1:topX,]
}

# Top links for particular parasite taxa
topPara <- function(P,Z,para,topX=10){
  rownames(P) <- rownames(Z)
  colnames(P) <- colnames(Z)  
  P_aux <- P[,colnames(P)%in%c(para)]
  Z_aux <- Z[,colnames(Z)%in%c(para)]
  P_aux[Z_aux==1]<--1
  aux = P_aux[order(P_aux,decreasing=TRUE)]
  aux <- data.frame(Host=names(aux), p=aux)
  rownames(aux) <- NULL
  colnames(aux)<-c('Host', 'p(interaction)')
    aux[1:topX,]  
}


# this should be in package but isn't loaded?
cross.validate.fold <-
function(Z, n= 10, min.per.col = 1, missing.pattern=c('random','prop.to.col.sums')){
    ## n-fold cross validation
    ## Returns a matrix of 3 columns, the first two are the (row,col) index of the pair,
    ## the third is the group
    missing.pattern = tolower(missing.pattern[1])
    if(max(range(Z))>1) Z[Z>0]<-1
    pairs = which(Z==1, arr.ind=T)
    colnames(pairs)<-c('row', 'col')
    
    if(length(which(colSums(Z)<min.per.col))>0){
        aux = which(pairs[,'col'] %in% which(colSums(Z)<min.per.col))
        if(length(aux))
            pairs = pairs[-aux,]
    }
    
    colm = pmax(colSums(Z) -min.per.col , 0)
    size = floor(sum(colm)/n)
    gr = rep(size, n)
    if(sum(colm) %% size!=0)
        gr[n] =  gr[n] + sum(colm) %% size
    
    group.colm = rep(1:n,times = gr)[sample.int(sum(colm), sum(colm))]
    pair.list = numeric(sum(colm))
    for(i in 1:sum(colm)){
        a = which(colm>0)
        if(missing.pattern=='random')
            b = a[sample.int(length(a),1)] else
        if (missing.pattern=='prop.to.col.sums')
            b = a[sample.int(length(a),1, prob=colm[a]/sum(colm[a]))] else
        stop('missing pattern has to be specified from selection!')
        colm[b] = colm[b]-1
        pair.list[i]<-b
    }
    pair.list= tapply(pair.list, group.colm,identity)
    
    gr.list= list()
    bank= c()
    for(i in 1:n){
        a= table(pair.list[[i]])
        gr.rows = unlist(sapply(1:length(a), function(r){
            b = which(pairs[,'col']== as.numeric(names(a[r])))
            b =setdiff(b, bank)
            b[sample.int(length(b), a[r])]
        }))
        bank = c(bank, gr.rows)
        gr.list[[i]]<-cbind(gr.rows, i)
    }

    aux = do.call('rbind', gr.list)
    pairs = cbind(pairs[aux[,1], ],gr= aux[,2])
    
    print(sprintf("Actual cross-validation rate is %0.3f" , table(pairs[,'gr'])/sum(1*(Z>0))))
    pairs[order(pairs[,'gr']),]
    
}

ana.table <-
function(Z, ZCross, P, roc, plot=FALSE){
    Z = 1*(Z>0)
    ZCross  = 1*(ZCross>0)
    Zpost = 1*(P>roc$threshold)
    data.frame(auc = roc$auc/100, thresh= roc$threshold,
               tot.ones = sum(Z), held.out.ones= sum(abs(ZCross - Z)[Z==1]),
               pred.held.out.ones = 100*sum(Zpost[Z==1 & ZCross==0])/sum(abs(ZCross - Z)[Z==1]),
               pred.tot.ones = sum(Zpost[Z==1])/sum(Z)*100)
    
}

```

# Data

Data downloaded April 29 2020 from VERENA google drive

```{r loadData, eval=TRUE, echo=TRUE}
cov_dat <- read.csv("../Data-20200429/BatCoV-assoc.csv", as.is=T)
tree_albery <- readRDS("../Data-20200429/Full Supertree.rds")
tree_super <- read.tree("../Data-20200429/supertree_mammals.tree")
# Greg's tree is not the Fritz super tree because there are 82 species missing from the data now, not 12 as with Greg's supertree.

# Dan Becker:
# "just need to change Artibeus phaeotis” and “Hipposideros commersoni” to “Dermanura phaeotis” and “Hipposideros commersonii” to match the names in “BatCoV-assoc.csv”"

# "off the top of my head, Perimyotis subflavus could be Pipistrellus subflavus, Carollia sowelli is probably Carollia brevicauda, Dermanura tolteca is probably Artibeus toltecus, Dermanura cinerea is probably Artibues cinera

# Artibeus planirostris has sometimes been considered a subspecies of A. jamaicensis

# Triaenops afer could be Triaenops persicus (same subspecies issue)

# Myotis pilosus is prolly M. ricketti"

# "Miniopterus fuliginosus was thought to be a subspecies of Miniopterus schreibersii"

# Chaerephon pusillus = Nyctinomus pusillus, Parastrellus hesperus = Pipistrellus hesperus, Neoromicia nana = Neoromicia nanus, Tadarida plicatus = Chaerephon plicatus

tree_albery$tip.label[tree_albery$tip.label=="Artibeus_phaeotis"] <- "Dermanura_phaeotis"
tree_albery$tip.label[tree_albery$tip.label=="Hipposideros_commersoni"] <- "Hipposideros_commersonii"
tree_albery$tip.label[tree_albery$tip.label=="Pipistrellus_subflavus"] <- "Perimyotis_subflavus"
tree_albery$tip.label[tree_albery$tip.label=="Carollia_brevicauda"] <- "Carollia_sowelli"
tree_albery$tip.label[tree_albery$tip.label=="Artibeus_toltecus"] <- "Dermanura_tolteca"

tree_albery$tip.label[tree_albery$tip.label=="Artibeus_cinera"] <- "Dermanura_cinerea"
# Artibeus_cinera not in tree

# tree_albery$tip.label[tree_albery$tip.label=="Myotis_ricketti"] <- "Myotis_pilosus"
# Can't do this because both myotis rickettii and pyotis pilosus are both in the data..

tree_albery$tip.label[tree_albery$tip.label=="Nyctinomus_pusillus"] <- "Chaerephon_pusillus"
# Nyctinomus_pusillus not in tree

tree_albery$tip.label[tree_albery$tip.label=="Pipistrellus_hesperus"] <- "Parastrellus_hesperus"
tree_albery$tip.label[tree_albery$tip.label=="Neoromicia_nanus"] <- "Neoromicia_nana"
tree_albery$tip.label[tree_albery$tip.label=="Chaerephon_plicatus"] <- "Tadarida_plicatus"

# Formatting host names to match phylogeny
cov_dat$host_species <- gsub(" ","_",cov_dat$host_species)

# Subspecies collapsed in data
cov_dat$host_species[cov_dat$host_species=="Artibeus_planirostris"] <- "Artibeus_jamaicensis"
cov_dat$host_species[cov_dat$host_species=="Triaenops_afer"] <- "Triaenops_persicus"
cov_dat$host_species[cov_dat$host_species=="Miniopterus_fuliginosus"] <- "Miniopterus_schreibersii"
cov_dat$host_species[cov_dat$host_species=="Myotis_pilosus"] <- "Myotis_ricketti"

# length(unique(cov_dat$host_species))#261
# length(tree_albery$tip.label)#5020
setdiff(cov_dat$host_species,tree_albery$tip.label) #4

# Creating binary interaction matrix 
com <- table(cov_dat$host_species, cov_dat$virus_genus)
com[com>1] <- 1
com <- as.matrix(unclass(com), nrow=nrow(com), ncol=ncol(com))

# testing: merge the trees and interaction matrices
dat <- network_clean(com, tree_albery, 'full')

####################################################################
# merging hp3 with BatCov at viral genus level
hp3 <- read.csv("../Data-20200429/HP3-associations.csv", as.is=T)

length(unique(hp3$hHostNameFinal))#754
length(intersect(hp3$hHostNameFinal,tree_albery$tip.label))#715
length(intersect(hp3$hHostNameFinal,tree_super$tip.label))#753
setdiff(hp3$hHostNameFinal,tree_super$tip.label)# Felis_concolor
tree_super$tip.label[grep("concolor",tree_super$tip.label)]
tree_super$tip.label["Puma_concolor"] <- "Felis_concolor"

# create hp3 with viruses set to genus level
virus_tax <- read.csv("../Data-20200429/HP3-viruses.csv", as.is=T)
hp3_genera <- left_join(hp3, virus_tax)
hp3_genera <- subset(hp3_genera, select=c(hHostNameFinal,vGenus,DetectionMethod,DetectionQuality))
names(hp3_genera)[1:2] <- c("host_species","virus_genus")

# join and remove double counted hp3 interactions
dat_full <- full_join(cov_dat[cov_dat$origin=="Anthony",], hp3_genera)
setdiff(dat_full$host_species,tree_super$tip.label)# 75
setdiff(dat_full$host_species,tree_albery$tip.label)# 40

com_full <- table(dat_full$host_species, dat_full$virus_genus)
com_full[com_full>1] <- 1
com_full <- as.matrix(unclass(com_full), nrow=nrow(com_full), ncol=ncol(com_full))

# merge the trees and interaction matrices
dat_full <- network_clean(com_full, tree_albery, 'full')
# tree_super doesn't seem to work well - multi2dierror

```


# Plotting interaction matrices

```{r plotZ_lof, eval=TRUE, echo=FALSE,fig.align='center'}
plot_Z(lof(dat$Z), tickMarks=1, cex.lab=1, cex.axis=1, ylab="host species", xlab="Viral genus - BatCov data")

plot_Z(lof(dat_full$Z), tickMarks=1, cex.lab=1, cex.axis=1, ylab="host species", xlab="Viral genus - merged BatCov + HP3")

```


# Models on BatCoV data

## Elmasri Phylogeny only model on BatCoV data

```{r batcov_phylo_cv_setup, echo=TRUE, eval=TRUE}
## General variables
MODEL = 'distance'                       # full, distance or affinity
SLICE = 1000                          # no of iterations
NO.CORES = 3                         # maximum cores to use
COUNT = TRUE                         # TRUE = count data, FALSE = year of first pub.
ALPHA.ROWS = 0.3					 # hyperparameter for prior over rows affinity, effective under affinity and full models only
ALPHA.COLS = 0.3					 # hyperparameter for prior over columns affinity, effective under affinity and full models only

## preparing tree and com
cleaned <- dat
com <- cleaned$Z                         # cleaned binary interaction matrix
tree <- cleaned$tree                     # cleaned tree
```

```{r batcov_phylo_cv, echo=TRUE, eval=FALSE}
## indexing 5-folds of interactions
folds <- cross.validate.fold(com, n = 10, min.per.col=1)  

# returns a matrix of 3 columns (row, col, group), (row, col) correspond to Z, group to the CV group
tot.gr <- length(unique(folds[,'gr']))   # total number of CV groups

## A loop to run over all CV groups
res <- mclapply(1:tot.gr, function(x, folds, Z, tree, slice, model.type, ALPHA.ROWS, ALPHA.COLS){

    ## Analysis for a single fold
    Z.train = Z
    Z.train[folds[which(folds[,'gr']==x),c('row', 'col')]]<-0

    ## running the model of interest
    obj = network_est(Z.train, slices=slice, tree=tree, model.type=model.type,
                      a_y = ALPHA.ROWS, a_w = ALPHA.COLS)

    P = sample_parameter(obj$param, model.type, Z.train, tree)
    Eta = if(is.null(obj$param$eta)) 0 else mean(obj$param$eta)
    
    ## order the rows in Z.test as in Z.train
    roc = rocCurves(Z, Z.train, P, plot=FALSE, bins=400, all=FALSE)
    tb  = ana.table(Z, Z.train, P, roc,  plot=FALSE)
    roc.all = rocCurves(Z, Z.train, P=P, plot=FALSE, bins=400, all=TRUE)
    tb.all  = ana.table(Z, Z.train, P, roc.all, plot=FALSE)
    
    list(param=list(P=P, Eta = Eta), tb = tb,
         tb.all = tb.all, FPR.all = roc.all$roc$FPR,
         TPR.all=roc.all$roc$TPR, FPR = roc$roc$FPR, TPR=roc$roc$TPR)
    
},  
	folds=folds, Z = com, tree=tree, model.type=MODEL, slice = SLICE,
    ALPHA.ROWS = ALPHA.ROWS, ALPHA.COLS= ALPHA.COLS, 
    mc.preschedule = TRUE, mc.cores = min(tot.gr, NO.CORES))

saveRDS(res,"../results/batcov_phylo_res.rds")

```

We can analyze the performance of the model via the area under the receiver operating characteristic curve (AUC), and the proportion of 1s in the original data successfully recovered. 

```{r batcov_phylo_diag, eval=TRUE, echo=FALSE}
res <- readRDS("../results/batcov_phylo_res.rds")

## Some analysis results, AUC, %1 recovered
TB = data.frame(
    m.auc = sapply(res, function(r) r$tb$auc),
    m.pred.held.out.ones = sapply(res,function(r) r$tb$pred.held.out.ones),
    m.thresh = sapply(res, function(r) r$tb$thresh),
    m.hold.out = sapply(res, function(r) r$tb$held.out.ones)
)
TB

## Printing and writing out average MCMC 
print(sprintf('Model: %s, AUC: %f and percent 1 recovered from held out: %f',
              MODEL,mean(TB$m.auc), mean(TB$m.pred.held.out.ones)))

## ROC curve points, can plot as plot(ROCgraph)
ROCgraph = cbind(
    FPR = rowMeans(sapply(res, function(r) r$FPR)),
    TPR = rowMeans(sapply(res, function(r) r$TPR)))

plot(ROCgraph, type="l", lty=2, col=2)

```

We construct the posterior probability matrix 'P' as the average across each fold, and look at the top undocumented interactions.

```{r batcov_phylo_topundoc, eval=TRUE, echo=TRUE}
## Constructing the P probability matrix from CV results
P = matrix(rowMeans(sapply(res, function(r) r$param$P)),
    nrow = nrow(cleaned$Z), ncol = ncol(cleaned$Z))

## view top undocumented interactions
topPairs(P,1*(cleaned$Z>0),topX=10)
topPara(P, 1*(cleaned$Z>0),topX=10, para="Betacoronavirus")

# Save full list of predictions
P[cleaned$Z==1]<--1 # set observed interactions to -1
rownames(P) <- rownames(cleaned$Z)
colnames(P) <- colnames(cleaned$Z)
aux = melt(P)
aux = aux[order(aux$value,decreasing=TRUE),]
colnames(aux)<-c('Host', 'Parasite', 'p(interaction)')

# remove observed interactions
aux <- aux[aux[,3]!=-1,]

# replace probabilities with ranks
aux$rank <- 1:nrow(aux)
aux <- aux[,-3]
write.csv(aux, "../results/batcov_elmasri_phylo_pred.csv")

aux_betacovs <- aux[aux$Parasite=="Betacoronavirus",]
# new ranks
aux_betacovs$rank <- 1:nrow(aux_betacovs)

write.csv(aux_betacovs, "../results/batcov_elmasri_phylo_pred_betacovsOnly.csv")

```


## Elmasri Full model on BatCoV data

```{r betacov_full_cv_setup, echo=TRUE, eval=TRUE}
## General variables
MODEL = 'full'                       # full, distance or affinity
SLICE = 1000                          # no of iterations
NO.CORES = 3                         # maximum cores to use
COUNT = TRUE                         # TRUE = count data, FALSE = year of first pub.
ALPHA.ROWS = 0.3                     # hyperparameter for prior over rows affinity, effective under affinity and full models only
ALPHA.COLS = 0.3                     # hyperparameter for prior over columns affinity, effective under affinity and full models only

## preparing tree and com
cleaned <- dat
com <- cleaned$Z                         # cleaned binary interaction matrix
tree <- cleaned$tree                     # cleaned tree

```

```{r betacov_full_cv, echo=TRUE, eval=FALSE}
## indexing 5-folds of interactions
folds <- cross.validate.fold(com, n = 10, min.per.col=1)  

# returns a matrix of 3 columns (row, col, group), (row, col) correspond to Z, group to the CV group
tot.gr <- length(unique(folds[,'gr']))   # total number of CV groups

## A loop to run over all CV groups
res <- mclapply(1:tot.gr, function(x, folds, Z, tree, slice, model.type, ALPHA.ROWS, ALPHA.COLS){

    ## Analysis for a single fold
    Z.train = Z
    Z.train[folds[which(folds[,'gr']==x),c('row', 'col')]]<-0

    ## running the model of interest
    obj = network_est(Z.train, slices=slice, tree=tree, model.type=model.type,
                      a_y = ALPHA.ROWS, a_w = ALPHA.COLS)

    P = sample_parameter(obj$param, model.type, Z.train, tree)
    Eta = if(is.null(obj$param$eta)) 0 else mean(obj$param$eta)
    
    ## order the rows in Z.test as in Z.train
    roc = rocCurves(Z, Z.train, P, plot=FALSE, bins=400, all=FALSE)
    tb  = ana.table(Z, Z.train, P, roc,  plot=FALSE)
    roc.all = rocCurves(Z, Z.train, P=P, plot=FALSE, bins=400, all=TRUE)
    tb.all  = ana.table(Z, Z.train, P, roc.all, plot=FALSE)
    
    list(param=list(P=P, Eta = Eta), tb = tb,
         tb.all = tb.all, FPR.all = roc.all$roc$FPR,
         TPR.all=roc.all$roc$TPR, FPR = roc$roc$FPR, TPR=roc$roc$TPR)
    
},  
    folds=folds, Z = com, tree=tree, model.type=MODEL, slice = SLICE,
    ALPHA.ROWS = ALPHA.ROWS, ALPHA.COLS= ALPHA.COLS, 
    mc.preschedule = TRUE, mc.cores = min(tot.gr, NO.CORES))

saveRDS(res,"../results/batcov_full_res.rds")

```

We can analyze the performance of the model via the area under the receiver operating characteristic curve (AUC), and the proportion of 1s in the original data successfully recovered. 

```{r batcov_full_diag, eval=TRUE, echo=FALSE}
res <- readRDS("../results/batcov_full_res.rds")

## Some analysis results, AUC, %1 recovered
TB = data.frame(
    m.auc = sapply(res, function(r) r$tb$auc),
    m.pred.held.out.ones = sapply(res,function(r) r$tb$pred.held.out.ones),
    m.thresh = sapply(res, function(r) r$tb$thresh),
    m.hold.out = sapply(res, function(r) r$tb$held.out.ones)
)
TB

## Printing and writing out average MCMC 
print(sprintf('Model: %s, AUC: %f and percent 1 recovered from held out: %f',
              MODEL,mean(TB$m.auc), mean(TB$m.pred.held.out.ones)))

## ROC curve points, can plot as plot(ROCgraph)
ROCgraph = cbind(
    FPR = rowMeans(sapply(res, function(r) r$FPR)),
    TPR = rowMeans(sapply(res, function(r) r$TPR)))

plot(ROCgraph, type="l", lty=2, col=2)

```

We construct the posterior probability matrix 'P' as the average across each fold, and look at the top undocumented interactions.

```{r batcov_full_topundoc, eval=TRUE, echo=TRUE}
## Constructing the P probability matrix from CV results
P = matrix(rowMeans(sapply(res, function(r) r$param$P)),
    nrow = nrow(cleaned$Z), ncol = ncol(cleaned$Z))

## view top undocumented interactions
topPairs(P,1*(cleaned$Z>0),topX=10)
topPara(P, 1*(cleaned$Z>0),topX=10, para="Betacoronavirus")

# Save full list of predictions
P[cleaned$Z==1]<--1 # set observed interactions to -1
rownames(P) <- rownames(cleaned$Z)
colnames(P) <- colnames(cleaned$Z)
aux = melt(P)
aux = aux[order(aux$value,decreasing=TRUE),]
colnames(aux)<-c('Host', 'Parasite', 'p(interaction)')

# remove observed interactions
aux <- aux[aux[,3]!=-1,]

# replace probabilities with ranks
aux$rank <- 1:nrow(aux)
aux <- aux[,-3]
write.csv(aux, "../results/batcov_elmasri_full_pred.csv")

aux_betacovs <- aux[aux$Parasite=="Betacoronavirus",]
# new ranks
aux_betacovs$rank <- 1:nrow(aux_betacovs)
write.csv(aux_betacovs, "../results/batcov_elmasri_full_pred_betacovsOnly.csv")

```


# Predicting with merged BatCov + HP3 (at viral genus level) 


## Elmasri Phylogeny model on merged data

```{r cv_phylo_hp3_setup, echo=TRUE, eval=FALSE}
## General variables
MODEL = 'distance'                       # full, distance or affinity
SLICE = 2000                          # no of iterations
NO.CORES = 3                         # maximum cores to use
COUNT = TRUE                         # TRUE = count data, FALSE = year of first pub.
ALPHA.ROWS = 0.3           # hyperparameter for prior over rows affinity, effective under affinity and full models only
ALPHA.COLS = 0.3           # hyperparameter for prior over columns affinity, effective under affinity and full models only

cleaned <- dat_full
com <- cleaned$Z                         # cleaned binary interaction matrix
tree <- cleaned$tree                     # cleaned tree
```

```{r cv_phylo_hp3, echo=TRUE, eval=FALSE}

## indexing 5-folds of interactions
folds <- cross.validate.fold(com, n = 10, min.per.col=1)  

# returns a matrix of 3 columns (row, col, group), (row, col) correspond to Z, group to the CV group
tot.gr <- length(unique(folds[,'gr']))   # total number of CV groups

## A loop to run over all CV groups
res <- mclapply(1:tot.gr, function(x, folds, Z, tree, slice, model.type, ALPHA.ROWS, ALPHA.COLS){

    ## Analysis for a single fold
    Z.train = Z
    Z.train[folds[which(folds[,'gr']==x),c('row', 'col')]]<-0

    ## running the model of interest
    obj = network_est(Z.train, slices=slice, tree=tree, model.type=model.type,
                      a_y = ALPHA.ROWS, a_w = ALPHA.COLS)

    P = sample_parameter(obj$param, model.type, Z.train, tree)
    Eta = if(is.null(obj$param$eta)) 0 else mean(obj$param$eta)
    
    ## order the rows in Z.test as in Z.train
    roc = rocCurves(Z, Z.train, P, plot=FALSE, bins=400, all=FALSE)
    tb  = ana.table(Z, Z.train, P, roc,  plot=FALSE)
    roc.all = rocCurves(Z, Z.train, P=P, plot=FALSE, bins=400, all=TRUE)
    tb.all  = ana.table(Z, Z.train, P, roc.all, plot=FALSE)
    
    list(param=list(P=P, Eta = Eta), tb = tb,
         tb.all = tb.all, FPR.all = roc.all$roc$FPR,
         TPR.all=roc.all$roc$TPR, FPR = roc$roc$FPR, TPR=roc$roc$TPR)
    
},  
  folds=folds, Z = com, tree=tree, model.type=MODEL, slice = SLICE,
    ALPHA.ROWS = ALPHA.ROWS, ALPHA.COLS= ALPHA.COLS, 
    mc.preschedule = TRUE, mc.cores = min(tot.gr, NO.CORES))

)

saveRDS(res,"../results/hp3_batcov_genera_phylo_res.rds")

```

We can analyze the performance of the model via the area under the receiver operating characteristic curve (AUC), and the proportion of 1s in the original data successfully recovered. 

```{r diagnostics, eval=TRUE, echo=FALSE}
res <- readRDS("../results/hp3_batcov_genera_phylo_res.rds")

## Some analysis results, AUC, %1 recovered
TB = data.frame(
    m.auc = sapply(res, function(r) r$tb$auc),
    m.pred.held.out.ones = sapply(res,function(r) r$tb$pred.held.out.ones),
    m.thresh = sapply(res, function(r) r$tb$thresh),
    m.hold.out = sapply(res, function(r) r$tb$held.out.ones)
)
TB

## Printing and writing out average MCMC 
print(sprintf('Model: %s, AUC: %f and percent 1 recovered from held out: %f',
              MODEL,mean(TB$m.auc), mean(TB$m.pred.held.out.ones)))

```

We construct the posterior probability matrix 'P' as the average across each fold, and look at the top undocumented interactions.

```{r hp3_phylo_topundoc_cv, eval=TRUE, echo=TRUE}
## Constructing the P probability matrix from CV results
P = matrix(rowMeans(sapply(res, function(r) r$param$P)),
    nrow = nrow(cleaned$Z), ncol = ncol(cleaned$Z))

## view top undocumented interactions
topPairs(P,1*(cleaned$Z>0),topX=10)
topPara(P, 1*(cleaned$Z>0),topX=10, para="Betacoronavirus")

# Save full list of predictions
P[cleaned$Z==1]<--1 # set observed interactions to -1
rownames(P) <- rownames(cleaned$Z)
colnames(P) <- colnames(cleaned$Z)
aux = melt(P)
aux = aux[order(aux$value,decreasing=TRUE),]
colnames(aux)<-c('Host', 'Parasite', 'p(interaction)')

# remove observed interactions
aux <- aux[aux[,3]!=-1,]

# replace probabilities with ranks
aux$rank <- 1:nrow(aux)
aux <- aux[,-3]
write.csv(aux, "../results/hp3_batcov_elmasri_phylo_pred.csv")

aux_betacovs <- aux[aux$Parasite=="Betacoronavirus",]
# new ranks
aux_betacovs$rank <- 1:nrow(aux_betacovs)
write.csv(aux_betacovs, "../results/hp3_batcov_elmasri_phylo_pred_betacovsOnly.csv")


```

## Elmasri Full model on merged data

```{r cv_full_hp3_setup, echo=TRUE, eval=TRUE}
## General variables
MODEL = 'full'                       # full, distance or affinity
SLICE = 2000                          # no of iterations
NO.CORES = 3                         # maximum cores to use
COUNT = TRUE                         # TRUE = count data, FALSE = year of first pub.
ALPHA.ROWS = 0.3           # hyperparameter for prior over rows affinity, effective under affinity and full models only
ALPHA.COLS = 0.3           # hyperparameter for prior over columns affinity, effective under affinity and full models only

cleaned <- dat_full
com <- cleaned$Z                         # cleaned binary interaction matrix
tree <- cleaned$tree                     # cleaned tree

```

```{r cv_full_hp3, echo=TRUE, eval=FALSE}
## indexing 5-folds of interactions
folds <- cross.validate.fold(com, n = 10, min.per.col=1)  

# returns a matrix of 3 columns (row, col, group), (row, col) correspond to Z, group to the CV group
tot.gr <- length(unique(folds[,'gr']))   # total number of CV groups

## A loop to run over all CV groups
res <- mclapply(1:tot.gr, function(x, folds, Z, tree, slice, model.type, ALPHA.ROWS, ALPHA.COLS){

    ## Analysis for a single fold
    Z.train = Z
    Z.train[folds[which(folds[,'gr']==x),c('row', 'col')]]<-0

    ## running the model of interest
    obj = network_est(Z.train, slices=slice, tree=tree, model.type=model.type,
                      a_y = ALPHA.ROWS, a_w = ALPHA.COLS)

    P = sample_parameter(obj$param, model.type, Z.train, tree)
    Eta = if(is.null(obj$param$eta)) 0 else mean(obj$param$eta)
    
    ## order the rows in Z.test as in Z.train
    roc = rocCurves(Z, Z.train, P, plot=FALSE, bins=400, all=FALSE)
    tb  = ana.table(Z, Z.train, P, roc,  plot=FALSE)
    roc.all = rocCurves(Z, Z.train, P=P, plot=FALSE, bins=400, all=TRUE)
    tb.all  = ana.table(Z, Z.train, P, roc.all, plot=FALSE)
    
    list(param=list(P=P, Eta = Eta), tb = tb,
         tb.all = tb.all, FPR.all = roc.all$roc$FPR,
         TPR.all=roc.all$roc$TPR, FPR = roc$roc$FPR, TPR=roc$roc$TPR)
    
},  
  folds=folds, Z = com, tree=tree, model.type=MODEL, slice = SLICE,
    ALPHA.ROWS = ALPHA.ROWS, ALPHA.COLS= ALPHA.COLS, 
    mc.preschedule = TRUE, mc.cores = min(tot.gr, NO.CORES))

)

saveRDS(res,"../results/hp3_batcov_genera_full_res.rds")

```

We can analyze the performance of the model via the area under the receiver operating characteristic curve (AUC), and the proportion of 1s in the original data successfully recovered. 

```{r full_hp3_diag, eval=TRUE, echo=FALSE}
res <- readRDS("../results/hp3_batcov_genera_full_res.rds")

## Some analysis results, AUC, %1 recovered
TB = data.frame(
    m.auc = sapply(res, function(r) r$tb$auc),
    m.pred.held.out.ones = sapply(res,function(r) r$tb$pred.held.out.ones),
    m.thresh = sapply(res, function(r) r$tb$thresh),
    m.hold.out = sapply(res, function(r) r$tb$held.out.ones)
)
TB

## Printing and writing out average MCMC 
print(sprintf('Model: %s, AUC: %f and percent 1 recovered from held out: %f',
              MODEL,mean(TB$m.auc), mean(TB$m.pred.held.out.ones)))

```

We construct the posterior probability matrix 'P' as the average across each fold, and look at the top undocumented interactions.

```{r topundoc_cv, eval=TRUE, echo=TRUE}
## Constructing the P probability matrix from CV results
P = matrix(rowMeans(sapply(res, function(r) r$param$P)),
    nrow = nrow(cleaned$Z), ncol = ncol(cleaned$Z))

## view top undocumented interactions
topPairs(P,1*(cleaned$Z>0),topX=10)
topPara(P, 1*(cleaned$Z>0),topX=10, para="Betacoronavirus")

# Save full list of predictions
P[cleaned$Z==1]<--1 # set observed interactions to -1
rownames(P) <- rownames(cleaned$Z)
colnames(P) <- colnames(cleaned$Z)
aux = melt(P)
aux = aux[order(aux$value,decreasing=TRUE),]
colnames(aux)<-c('Host', 'Parasite', 'p(interaction)')

# remove observed interactions
aux <- aux[aux[,3]!=-1,]

# replace probabilities with ranks
aux$rank <- 1:nrow(aux)
aux <- aux[,-3]
write.csv(aux, "../results/hp3_batcov_elmasri_full_pred.csv")

aux_betacovs <- aux[aux$Parasite=="Betacoronavirus",]
# new ranks
aux_betacovs$rank <- 1:nrow(aux_betacovs)
write.csv(aux_betacovs, "../results/hp3_batcov_elmasri_full_pred_betacovsOnly.csv")


```



## References

Elmasri, M., Farrell, M. J., Davies, T. J., & Stephens, D. A. (2020). A hierarchical Bayesian model for predicting ecological interactions using scaled evolutionary relationships. [Annals of Applied Statistics, 14(1), 221-240.](https://projecteuclid.org/euclid.aoas/1587002672)


