---
title: "Bat-Betacoronavirus link prediction via Elmasri et al. approach"
author: "Maxwell J. Farrell"
date: "`r format(Sys.time(), '%B %d %Y')`"
output: 
  html_document:
    highlight: default
    toc: yes
    toc_depth: 2
    toc_float: true
    theme: yeti
urlcolor: blue
---

```{r setup, include = FALSE, eval = TRUE, echo = TRUE, message = FALSE}

knitr::opts_chunk$set(
  
  collapse = TRUE,
  
  comment = "#>"
  
)

```

```{r loadinglibraries}
# devtools::install_github("melmasri/HPprediction")
require(HPprediction)
require(ape)
require(dplyr)
require(parallel)

conflict_prefer("rescale", "geiger")

```

```{r custom_funcs, eval = T, echo = F}
# Custom functions to parse predictions

# Modified topPairs function that attaches names from Z matrix
topPairs <- function(P,Z,topX = 20){
  ## Plotting the top uncodumented interactions
  P[Z == 1] <- -1
  rownames(P) <- rownames(Z)
  colnames(P) <- colnames(Z)
  aux = reshape2::melt(P)
  aux = aux[order(aux$value,decreasing = TRUE),]
  colnames(aux) <- c('Host', 'Parasite', 'p(interaction)')
  aux
  aux[1:topX,]
}

# Top links for particular parasite taxa
topPara <- function(P,Z,para,topX = 10){
  rownames(P) <- rownames(Z)
  colnames(P) <- colnames(Z)  
  P_aux <- P[,colnames(P)%in%c(para)]
  Z_aux <- Z[,colnames(Z)%in%c(para)]
  P_aux[Z_aux == 1] <- -1
  aux = P_aux[order(P_aux,decreasing = TRUE)]
  aux <- data.frame(Host = names(aux), p = aux)
  rownames(aux) <- NULL
  colnames(aux) <- c('Host', 'p(interaction)')
  aux[1:topX,]  
}


# this should be in package but isn't loaded?
cross.validate.fold  <- 
  function(Z, n =  10, min.per.col = 1, missing.pattern = c('random','prop.to.col.sums')){
    ## n-fold cross validation
    ## Returns a matrix of 3 columns, the first two are the (row,col) index of the pair,
    ## the third is the group
    missing.pattern = tolower(missing.pattern[1])
    if(max(range(Z))>1) Z[Z>0] <- 1
    pairs = which(Z == 1, arr.ind = T)
    colnames(pairs) <- c('row', 'col')
    
    if(length(which(colSums(Z)<min.per.col))>0){
      aux = which(pairs[,'col'] %in% which(colSums(Z)<min.per.col))
      if(length(aux))
        pairs = pairs[-aux,]
    }
    
    colm = pmax(colSums(Z) -min.per.col , 0)
    size = floor(sum(colm)/n)
    gr = rep(size, n)
    if(sum(colm) %% size!= 0)
      gr[n] =  gr[n] + sum(colm) %% size
    
    group.colm = rep(1:n,times = gr)[sample.int(sum(colm), sum(colm))]
    pair.list = numeric(sum(colm))
    for(i in 1:sum(colm)){
      a = which(colm>0)
      if(missing.pattern == 'random')
        b = a[sample.int(length(a),1)] else
          if (missing.pattern == 'prop.to.col.sums')
            b = a[sample.int(length(a),1, prob = colm[a]/sum(colm[a]))] else
              stop('missing pattern has to be specified from selection!')
          colm[b] = colm[b]-1
          pair.list[i] <- b
    }
    pair.list =  tapply(pair.list, group.colm,identity)
    
    gr.list =  list()
    bank =  c()
    for(i in 1:n){
      a =  table(pair.list[[i]])
      gr.rows = unlist(sapply(1:length(a), function(r){
        b = which(pairs[,'col'] ==  as.numeric(names(a[r])))
        b  = setdiff(b, bank)
        b[sample.int(length(b), a[r])]
      }))
      bank = c(bank, gr.rows)
      gr.list[[i]] <- cbind(gr.rows, i)
    }
    
    aux = do.call('rbind', gr.list)
    pairs = cbind(pairs[aux[,1], ],gr =  aux[,2])
    
    print(sprintf("Actual cross-validation rate is %0.3f" , table(pairs[,'gr'])/sum(1*(Z>0))))
    pairs[order(pairs[,'gr']),]
    
  }

ana.table  <- 
  function(Z, ZCross, P, roc, plot = FALSE){
    Z = 1*(Z>0)
    ZCross  = 1*(ZCross>0)
    Zpost = 1*(P>roc$threshold)
    data.frame(auc = roc$auc/100, thresh =  roc$threshold,
               tot.ones = sum(Z), held.out.ones =  sum(abs(ZCross - Z)[Z == 1]),
               pred.held.out.ones = 100*sum(Zpost[Z == 1 & ZCross == 0])/sum(abs(ZCross - Z)[Z == 1]),
               pred.tot.ones = sum(Zpost[Z == 1])/sum(Z)*100)
    
  }

```

# Data

## Loading 

```{r loadData, eval = TRUE, echo = TRUE}

# Data from virionette pulled May 13th 2020 (commit 33f56ad)

bat_tree <- readRDS(paste0(here::here(), 
                           "/Github/Repos/virionette/04_predictors/bat-supertree_clean.rds"))

dat <- read.csv(paste0(here::here(), 
                       "/Github/Repos/virionette/03_interaction_data/virionette.csv"), as.is = T)

# Subset to Chiroptera
dat <- dat[dat$host_order == "Chiroptera",]

# format host names to match tree
dat$host_species <- gsub(" ","_", dat$host_species)
setdiff(dat$host_species,bat_tree$tip.label)# 0

com <- table(dat$host_species, dat$virus_genus)
com[com>1] <- 1
com <- as.matrix(unclass(com), nrow = nrow(com), ncol = ncol(com))

# merge the trees and interaction matrices
dat_bats <- network_clean(com, bat_tree, 'full')
dim(dat_bats$Z)# 254 x 14


```

# Plotting interaction matrix

```{r plotZ_lof, eval = TRUE, echo = FALSE,fig.align = 'center'}
plot_Z(lof(dat_bats$Z), tickMarks = 1, cex.lab = 1, cex.axis = 1, ylab = "Bat host species", xlab = "Viral genus ")

```


# Elmasri combined full model

```{r betacov_full_cv_setup, echo = TRUE, eval = TRUE}
## General variables
MODEL = 'full'                       # full, distance or affinity
SLICE = 1000                          # no of iterations
NO.CORES = 1                        # maximum cores to use
COUNT = TRUE                         # TRUE = count data, FALSE = year of first pub.
ALPHA.ROWS = 0.3                     # hyperparameter for prior over rows affinity, effective under affinity and full models only
ALPHA.COLS = 0.3                     # hyperparameter for prior over columns affinity, effective under affinity and full models only

## preparing tree and com
cleaned <- dat_bats
com <- cleaned$Z                         # cleaned binary interaction matrix
tree <- cleaned$tree                     # cleaned tree

```

```{r betacov_full_cv, echo = TRUE, eval = TRUE}

if(!file.exists("results/bats_viral_genera_full_res.rds")){
  
  ## indexing 10-folds of interactions
  folds <- cross.validate.fold(com, n = 10, min.per.col = 2)  
  
  # returns a matrix of 3 columns (row, col, group), (row, col) correspond to Z, group to the CV group
  tot.gr <- length(unique(folds[,'gr']))   # total number of CV groups
  
  ## A loop to run over all CV groups
  res <- mclapply(1:tot.gr, function(x, folds, Z, tree, slice, model.type, ALPHA.ROWS, ALPHA.COLS){
    
    ## Analysis for a single fold
    Z.train = Z
    Z.train[folds[which(folds[,'gr'] == x),c('row', 'col')]] <- 0
    
    ## running the model of interest
    obj = network_est(Z.train, slices = slice, tree = tree, model.type = model.type,
                      a_y = ALPHA.ROWS, a_w = ALPHA.COLS)
    
    P = sample_parameter(obj$param, model.type, Z.train, tree)
    Eta = if(is.null(obj$param$eta)) 0 else mean(obj$param$eta)
    
    ## order the rows in Z.test as in Z.train
    roc = rocCurves(Z, Z.train, P, plot = FALSE, bins = 400, all = FALSE)
    tb  = ana.table(Z, Z.train, P, roc,  plot = FALSE)
    roc.all = rocCurves(Z, Z.train, P = P, plot = FALSE, bins = 400, all = TRUE)
    tb.all  = ana.table(Z, Z.train, P, roc.all, plot = FALSE)
    
    list(param = list(P = P, Eta = Eta), tb = tb,
         tb.all = tb.all, FPR.all = roc.all$roc$FPR,
         TPR.all = roc.all$roc$TPR, FPR = roc$roc$FPR, TPR = roc$roc$TPR)
    
  },  
  folds = folds, Z = com, tree = tree, model.type = MODEL, slice = SLICE,
  ALPHA.ROWS = ALPHA.ROWS, ALPHA.COLS =  ALPHA.COLS, 
  mc.preschedule = TRUE, mc.cores = min(tot.gr, NO.CORES))
  
  saveRDS(res, "results/bats_viral_genera_full_res.rds")
  
}

```

We can analyze the performance of the model via the area under the receiver operating characteristic curve (AUC), and the proportion of 1s in the original data successfully recovered. 

```{r batcov_full_diag, eval = TRUE, echo = FALSE}

res <- readRDS("results/bats_viral_genera_full_res.rds")

## Some analysis results, AUC, %1 recovered
TB = data.frame(
  m.auc = sapply(res, function(r) r$tb$auc),
  m.pred.held.out.ones = sapply(res,function(r) r$tb$pred.held.out.ones),
  m.thresh = sapply(res, function(r) r$tb$thresh),
  m.hold.out = sapply(res, function(r) r$tb$held.out.ones)
)
TB

## Printing and writing out average MCMC 
print(sprintf('Model: %s, AUC: %f and percent 1 recovered from held out: %f',
              MODEL,mean(TB$m.auc), mean(TB$m.pred.held.out.ones)))

## ROC curve points, can plot as plot(ROCgraph)
ROCgraph = cbind(
  FPR = rowMeans(sapply(res, function(r) r$FPR)),
  TPR = rowMeans(sapply(res, function(r) r$TPR)))

plot(ROCgraph, type = "l", lty = 2, col = 2)

```

## Plot of phylogeny scaling

```{r eta_plots, eval = TRUE, echo = FALSE}

etas <- sapply(res, function(r) r$param$Eta)
eta <- mean(etas)

png("results/eta_plot.png")
par(mfrow = c(1,2))
plot(tree, show.tip.label = F, no.margin = FALSE)
title("Unscaled tree ")
plot(rescale(tree, model = "EB", a = eta), show.tip.label = F)
title(paste0("Scaled tree ","(eta = ",round(eta,2),")"))
dev.off()


```

We construct the posterior probability matrix 'P' as the average across each fold, and look at the top undocumented interactions.

```{r batcov_full_topundoc, eval = TRUE, echo = TRUE}
## Constructing the P probability matrix from CV results
P = matrix(rowMeans(sapply(res, function(r) r$param$P)),
           nrow = nrow(cleaned$Z), ncol = ncol(cleaned$Z))

## view top undocumented interactions
topPairs(P,1*(cleaned$Z>0),topX = 10)
topPara(P, 1*(cleaned$Z>0),topX = 10, para = "Betacoronavirus")

# Save full list of predictions
# P[cleaned$Z == 1] <- -1 # set observed interactions to -1
rownames(P) <- rownames(cleaned$Z)
colnames(P) <- colnames(cleaned$Z)
aux = reshape2::melt(P)
aux = aux[order(aux$value,decreasing = TRUE),]
colnames(aux) <- c('Host', 'Parasite', 'p.interaction')

# Subset to betacovs
aux <- aux[aux$Parasite == "Betacoronavirus",]

# adding bats in the tree but without data to make predictions
bats_nodata <- data.frame(Host = setdiff(bat_tree$tip.label,rownames(com)),
                          Parasite = NA,
                          'p.interaction' = NA)

aux <- rbind(aux, bats_nodata)

# remove Parasite column
aux <- select(aux, -Parasite)

write.csv(aux, "results/FarrellBatFull.csv", row.names = FALSE)

```



# Elmasri phylogeny only model

```{r betacov_phylo_cv_setup, echo = TRUE, eval = TRUE}
## General variables
MODEL = 'distance'                       # full, distance or affinity
SLICE = 1000                          # no of iterations
NO.CORES = 1                       # maximum cores to use
COUNT = TRUE                         # TRUE = count data, FALSE = year of first pub.
ALPHA.ROWS = 0.3                     # hyperparameter for prior over rows affinity, effective under affinity and full models only
ALPHA.COLS = 0.3                     # hyperparameter for prior over columns affinity, effective under affinity and full models only

## preparing tree and com
cleaned <- dat_bats
com <- cleaned$Z                         # cleaned binary interaction matrix
tree <- cleaned$tree                     # cleaned tree

```


```{r betacov_phylo_cv, echo = TRUE, eval = TRUE}

if(!file.exists("results/bats_viral_genera_phylo_res.rds")){
  
  ## indexing 10-folds of interactions
  folds <- cross.validate.fold(com, n = 10, min.per.col = 2)  
  
  # returns a matrix of 3 columns (row, col, group), (row, col) correspond to Z, group to the CV group
  tot.gr <- length(unique(folds[,'gr']))   # total number of CV groups
  
  ## A loop to run over all CV groups
  res <- mclapply(1:tot.gr, function(x, folds, Z, tree, slice, model.type, ALPHA.ROWS, ALPHA.COLS){
    
    ## Analysis for a single fold
    Z.train = Z
    Z.train[folds[which(folds[,'gr'] == x),c('row', 'col')]] <- 0
    
    ## running the model of interest
    obj = network_est(Z.train, slices = slice, tree = tree, model.type = model.type,
                      a_y = ALPHA.ROWS, a_w = ALPHA.COLS)
    
    P = sample_parameter(obj$param, model.type, Z.train, tree)
    Eta = if(is.null(obj$param$eta)) 0 else mean(obj$param$eta)
    
    ## order the rows in Z.test as in Z.train
    roc = rocCurves(Z, Z.train, P, plot = FALSE, bins = 400, all = FALSE)
    tb  = ana.table(Z, Z.train, P, roc,  plot = FALSE)
    roc.all = rocCurves(Z, Z.train, P = P, plot = FALSE, bins = 400, all = TRUE)
    tb.all  = ana.table(Z, Z.train, P, roc.all, plot = FALSE)
    
    list(param = list(P = P, Eta = Eta), tb = tb,
         tb.all = tb.all, FPR.all = roc.all$roc$FPR,
         TPR.all = roc.all$roc$TPR, FPR = roc$roc$FPR, TPR = roc$roc$TPR)
    
  },  
  folds = folds, Z = com, tree = tree, model.type = MODEL, slice = SLICE,
  ALPHA.ROWS = ALPHA.ROWS, ALPHA.COLS =  ALPHA.COLS, 
  mc.preschedule = TRUE, mc.cores = min(tot.gr, NO.CORES))
  
  saveRDS(res, "results/bats_viral_genera_phylo_res.rds")
  
}

```

We can analyze the performance of the model via the area under the receiver operating characteristic curve (AUC), and the proportion of 1s in the original data successfully recovered. 

```{r batcov_phylo_diag, eval = TRUE, echo = FALSE}

res <- readRDS("results/bats_viral_genera_phylo_res.rds")

## Some analysis results, AUC, %1 recovered
TB = data.frame(
  m.auc = sapply(res, function(r) r$tb$auc),
  m.pred.held.out.ones = sapply(res,function(r) r$tb$pred.held.out.ones),
  m.thresh = sapply(res, function(r) r$tb$thresh),
  m.hold.out = sapply(res, function(r) r$tb$held.out.ones)
)
TB

## Printing and writing out average MCMC 
print(sprintf('Model: %s, AUC: %f and percent 1 recovered from held out: %f',
              MODEL,mean(TB$m.auc), mean(TB$m.pred.held.out.ones)))

## ROC curve points, can plot as plot(ROCgraph)
ROCgraph = cbind(
  FPR = rowMeans(sapply(res, function(r) r$FPR)),
  TPR = rowMeans(sapply(res, function(r) r$TPR)))

plot(ROCgraph, type = "l", lty = 2, col = 2)

```

## Plot of phylogeny scaling

```{r eta_plots_2, eval = TRUE, echo = FALSE}

etas <- sapply(res, function(r) r$param$Eta)
eta <- mean(etas)

png("results/eta_plot_phylo.png")
par(mfrow = c(1,2))
plot(tree, show.tip.label = F, no.margin = FALSE)
title("Unscaled tree ")
plot(rescale(tree, model = "EB", a = eta), show.tip.label = F)
title(paste0("Scaled tree ","(eta = ",round(eta,2),")"))
dev.off()


```

We construct the posterior probability matrix 'P' as the average across each fold, and look at the top undocumented interactions.

```{r batcov_phylo_topundoc, eval = TRUE, echo = TRUE}
## Constructing the P probability matrix from CV results
P = matrix(rowMeans(sapply(res, function(r) r$param$P)),
           nrow = nrow(cleaned$Z), ncol = ncol(cleaned$Z))

## view top undocumented interactions
topPairs(P,1*(cleaned$Z>0),topX = 10)
topPara(P, 1*(cleaned$Z>0),topX = 10, para = "Betacoronavirus")

# Save full list of predictions
# P[cleaned$Z == 1] <- -1 # set observed interactions to -1
rownames(P) <- rownames(cleaned$Z)
colnames(P) <- colnames(cleaned$Z)
aux = reshape2::melt(P)
aux = aux[order(aux$value,decreasing = TRUE),]
colnames(aux) <- c('Host', 'Parasite', 'p.interaction')

# Setting SHP predictions to NA 
# will be fixed in next HP Predict commit
# names(which(colSums(cleaned$Z) == 1))
aux$p.interaction[aux$p.interaction == 1] <- NA

# Subset to betacovs
aux <- aux[aux$Parasite == "Betacoronavirus",]

# adding bats in the tree but without data to make predictions
bats_nodata <- data.frame(Host = setdiff(bat_tree$tip.label,rownames(com)),
                          Parasite = NA,
                          'p.interaction' = NA)

aux <- rbind(aux, bats_nodata)

# remove Parasite column
aux <- select(aux, -Parasite)

write.csv(aux, "results/FarrellBatPhylogeny.csv", row.names = FALSE)

```
