---
title: "Plug and play link estimation"
author: "Tad Dallas"
output:
  pdf_document:
    toc: yes
  html_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: tango
    theme: journal
    toc: yes
---





```{r eval = TRUE, echo = FALSE, comment = "", message = FALSE}

#dat <- read.csv(paste0(here::here(), "/Github/Repos/virionette/03_interaction_data/virionette.csv"), stringsAsFactors = FALSE)

#host <- read.csv(paste0(here::here(), "/Github/Repos/virionette/04_predictors/Han-BatTraits.csv"), stringsAsFactors = FALSE)

#cites <- read.csv(paste0(here::here(), "/Github/Repos/virionette/04_predictors/Citations.csv"), stringsAsFactors = FALSE)

# dat <- read.delim('models/virionette.txt', stringsAsFactors=FALSE, sep = ",")
#dat <- read.csv('models/virionette.csv', stringsAsFactors=FALSE)
#host <- read.csv('models/Han-BatTraits.csv',stringsAsFactors=FALSE)
#cites <- read.csv('models/Citations.csv',stringsAsFactors=FALSE)

#dat <- read.csv('virionette.csv', stringsAsFactors=FALSE)
#host <- read.csv('Han-BatTraits.csv',stringsAsFactors=FALSE)
#cites <- read.csv('Citations.csv',stringsAsFactors=FALSE)

dat <- read.csv(paste0("C:/Users/gfalb/Documents/Script/Fresnel/", "/Github/Repos/virionette/03_interaction_data/virionette.csv"), stringsAsFactors = FALSE)

host <- read.csv(paste0("C:/Users/gfalb/Documents/Script/Fresnel/", "/Github/Repos/virionette/04_predictors/Han-BatTraits.csv"), stringsAsFactors = FALSE)

cites <- read.csv(paste0("C:/Users/gfalb/Documents/Script/Fresnel/", "/Github/Repos/virionette/04_predictors/Citations.csv"), stringsAsFactors = FALSE)

# set.seed(666)

```




# Load packages

```{r}

library(Amelia)
library(fields)
library(ROCR)
library(corpcor)
library(np)
library(plyr)
library(dplyr)
library(vegan)

```



# Make some colors 
```{r}
ose1 <- rgb(85,108,17, m = 255)
ose2 <- rgb(160,108,17, m = 255)
ose3 <- rgb(114,132,56, m = 255)
ose4 <- rgb(137,152,87, m = 255)
```





```{r, eval = TRUE, echo = FALSE}

#' Plug and Play algorithm
#' Applies the plug and play algorithm to data on environmental covariates at presence and background using multivariate Gaussian distribution for f and f1
#'
#'  @param p covariate data at presence/occurrence points
#'  @param bgrd covariate data at background points
#'  @param method either 'classical' (default), 'robust', or 'regularized'
#'  
#'  @returns A list contain the means and covariances
#'  

pp.gauss <- function(p, bgrd, method = 'regularized', vbose = FALSE){
  mean.p <- colMeans(p)         # estimated mean of presence points
  mean.bgrd <- colMeans(bgrd)   # estimated mean of background points
  sigma.p <- stats::cov(p)             # estimated covariance of presence points
  sigma.bgrd <- stats::cov(bgrd)       # estimated covariance of background points
  
  if(method == 'robust'){
    require(robust)
    p.est <- covRob(p)
    mean.p <- p.est$center        # robust estimated mean
    sigma.p <- p.est$cov          # robust estimated covariance of presence points
    bgrd.est <- covRob(bgrd)
    mean.bgrd <- bgrd.est$center  # robust estimated mean
    sigma.bgrd <- bgrd.est$cov    # robust estimated covariance of background points
  }
  if(method == 'regularized'){
    require(corpcor)
    # robust estimated covariance of presence points
    p.est <- as.numeric(cov.shrink(p, verbose =  vbose))  
    sigma.p <- matrix(as.numeric(p.est), nrow = sqrt(length(p.est)))  
    # robust estimated covariance of background points
    bgrd.est <- as.numeric(cov.shrink(bgrd, verbose = vbose))    
    sigma.bgrd <- matrix(as.numeric(bgrd.est), nrow = sqrt(length(bgrd.est))) # reformat
  }
  return(model = list(mean.p = mean.p, sigma.p = sigma.p, mean.bgrd = mean.bgrd, sigma.bgrd = sigma.bgrd))
}



#'  Takes a learned gaussian-plug-and-play and applies to a new point
#'  
#'  @param model a list object from pp.gauss
#'  @param x  a vector of points to classify
#'
#'  @returns A vector of relative suitabilities

predict.pp.gauss <- function(model, x){
  require(mvtnorm)
  f1 <- dmvnorm(x, mean = model$mean.p, sigma = model$sigma.p, log = TRUE)
  f0 <- dmvnorm(x, mean = model$mean.bgrd, sigma = model$sigma.bgrd, log = TRUE)
  return(S = exp(f1-f0))
}




#' Applies the plug and play algorithm to data on environmental covariates at presence and background using multivariate Gaussian distribution for f and f1
#'
#'  @param p covariate data at presence/occurrence points
#'  @param bgrd covariate data at background points
#'
#'  @returns model a list object containing f1 and f0 

pp.kde <- function(p, bgrd){
  require(np)
  f1 <- npudens(p, bwmethod = 'normal-reference')
  f0 <- npudens(bgrd, bwmethod = 'normal-reference')
  return(model = list(f1 = f1,f0 = f0))
}




#'  Takes a learned kde-plug-and-play and applies to a new point
#'  
#'  @param model a list object from pp.gauss
#'  @param x  a vector of points to classify
#'
#'  @returns A vector of relative suitabilities

predict.pp.kde <- function(model, x){
  require(np)
  f1 <- fitted(npudens(bws = model$f1$bws, edat = data.frame(x), bwmethod = 'normal-reference'))
  f0 <- fitted(npudens(bws = model$f0$bws, edat = data.frame(x), bwmethod = 'normal-reference'))
  return(S = f1/f0)
}

```


# The code below is used to calculate the number of estimated missing links from host-parasite networks


```{r eval = TRUE, echo = F}

#' Function to compute frequnecy of frequencies table from a list of records
#' Input: x (a list of unique records)
#' 
#'
#' @returns frequency of frequencies table 

FoF <- function(x){
  x.0 <- table(table(x))
  out <- cbind(as.numeric(names(x.0)),as.vector(x.0))
  out
}



#' Function to estimate and summarize missing classes
#' @param fof (a frequency-of-frequencies table)
#' 
#' @returns list object containing observed, estimated, and missing links

summarize <- function(fof){
  observed <- sum(fof[,2])
  ace <- ChaoLee1992(fof, t = 10, method = "all", conf = 0.95)
  number.missing <- ace$Nhat - observed
  fraction.observed <- observed/ace$Nhat[2:1]
  out <- list(observed = observed, estimated = ace$Nhat[2:1], number.mising = number.missing, fraction.observed = fraction.observed)
  out
}

```






---



# Host-virus associations


```{r eval = TRUE, echo = TRUE}

#' imputes missing data.
#'
#' @param listobject contains interaction matrix (intMat), host traits, parasite traits, and a points object that gets handed to pp.gauss
#' 
#' @returns the same listobject, now with missing host and parasite trait data imputed and pointsobject updated.

getImpute <- function(listobject, m = 10, rem = FALSE, useMeans = TRUE){
  require(Amelia)
  require(vegan)
  
  htriz <- listobject$Hosts
  ptriz <- listobject$Parasites
  
  #remove columns with less than 75% data coverage
  no  <- which(colSums(is.na(htriz))/nrow(htriz) > 0.75)
  nop <- which(colSums(is.na(ptriz))/nrow(ptriz) > 0.75)
  if(length(no) > 0){
    htriz <- htriz[,-no]
  }
  if(length(nop) > 0){
    ptriz <- ptriz[,-nop]
  }
  
  # remove columns with correlation coeff
  no2 <- which(cor(htriz, use = 'complete.obs')[,1] > 0.6)
  if(length(no2) > 1){
    no2 <- no2[-1]
    htriz = htriz[,-no2]
  }
  
  #remove rows without any data
  #hosts
  inds <- which(is.na(colSums(htriz)))
  if(rem == TRUE){
    cut = rep(0,nrow(htriz))
    for(i in seq_len(nrow(htriz))){
      if(all(is.na(htriz[i,]))){cut[i] = 1}
    }
    htriz <- htriz[-which(cut == 1), ]
  }
  #parasites
  indsP <- which(is.na(colSums(ptriz)))
  if(rem == TRUE){
    cut = rep(0,nrow(ptriz))
    for(i in seq_len(nrow(ptriz))){
      if(all(is.na(ptriz[i,]))){cut[i] = 1}
    }
    ptriz <- ptriz[-which(cut == 1), ]
  }
  
  #impute host traits
  if(any(is.na(colSums(htriz)))){
    helper <- which(colSums(htriz,na.rm = TRUE)>1)[1]
    tmp <- htriz[, unique(c(helper,inds))]
    if(useMeans == TRUE){
      means <- which(apply(tmp, 1, function(x){sum(is.na(x))})/ncol(tmp) > 0)
      tmp[means, ] <- colMeans(tmp, na.rm = TRUE)
      htriz2 <- tmp
    }else{
      htriz2 <- amelia(x = tmp, m = m, p2s = 0)$imputations[[m]]
    }
    notInds <- seq(1,ncol(htriz),by = 1)[-c(helper,inds)]
    htriz3 <- cbind(htriz2, htriz[,notInds])
  }else{
    htriz2 <- amelia(x = htriz[, c(inds)], m = m, p2s = 0)$imputations[[m]]
    notInds <- seq(1,ncol(htriz),by = 1)[-c(inds)]
    htriz3 <- cbind(htriz2, htriz[,notInds])
  }
  
  
  #remake the points object
  intMat <- listobject$intMat
  pos <- which(intMat  ==  1, arr.ind = TRUE)
  neg <- which(intMat  ==  0, arr.ind = TRUE)
  ret <- rbind(pos,neg) 
  ret <- cbind(ret, c(rep(1, nrow(pos)), rep(0,nrow(neg)))) 
  colnames(ret) <- c('parasite', 'host', 'presence')
  
  pt <- ptriz[ret[,2],]
  ht <- htriz3[ret[,1],]
  ht <- ht[,-1]
  
  ptobj <- data.frame(ret, ht, pt, stringsAsFactors = TRUE)
  return(list(intMat = intMat, Hosts = htriz3, 
              Parasites = ptriz, pointsObject = ptobj))
}


```


```{r eval = TRUE, echo = TRUE}

#' Train pp.gauss model on data
#'
#' @param pointsobject input to pp.gauss function
#' 
#' @returns list object of host and parasite information, model predictions, and AUC value

getReal <- function(pointsobject, cv = TRUE, folds = 5){
  require(ROCR)
  # all the data divided into presence (p) and background (bg)
  p <- pointsobject[(pointsobject[,'presence']  ==  1), 4:ncol(pointsobject)]
  bg <- pointsobject[,4:ncol(pointsobject)]
  if(cv == TRUE){
    # Take a random sample and exclude it
    ints <- sample(1:nrow(pointsobject), nrow(pointsobject)/folds)
    red <- pointsobject[-ints,]
    p.red <- red[(red[,'presence']  ==  1), 4:ncol(pointsobject)]
    bg.red <- red[, 4:ncol(pointsobject)]
    # Train the model on a portion of the data
    cvmodel <- pp.gauss(p = p.red, bgrd = bg.red, vbose = FALSE)
    #... and predict on the holdout set
    red.preds <- predict.pp.gauss(cvmodel, x = bg[ints,])
    ret <- data.frame(presence = pointsobject[ints,'presence'], suitability = red.preds)
    hosts <- pointsobject[ints,'parasite']; parasites = pointsobject[ints,'host']
  }
  
  if(cv == FALSE){
    fullmodel <- pp.gauss(p = p, bgrd = bg, vbose = FALSE)
    # ignore cites in the predictions on bg data
    if(!is.null(bg$cites)){
      bg$cites <- mean(bg$cites)
    }
    redpreds <- predict.pp.gauss(fullmodel, x = bg)
    ret <- data.frame(presence = pointsobject[,'presence'], suitability = redpreds)
    hosts <- pointsobject$parasite; parasites = pointsobject$host
  }
  #auc calculation
  pred <- prediction(predictions = ret[,2], labels = ret[,1])
  acc <- as.numeric(performance(pred, 'auc')@y.values)
  return(list(hostInds = hosts, parasiteInds = parasites, predictions = ret, auc = acc))
}

```






```{r eval = TRUE, echo = F, message = FALSE, warning = FALSE}

makePoints <- function(intmat, HostTraits, ParasiteTraits){
  ret <- matrix(0,ncol = 3, nrow = (nrow(intmat)*ncol(intmat))) 
  colnames(ret) <- c('row', 'column', 'presence')
  negInts <- which(intmat  ==  0, arr.ind = TRUE)
  posInts <- which(intmat > 0, arr.ind = TRUE)
  ret[,1:2] <- rbind(negInts, posInts)
  ret[,3] <- c(rep(0, nrow(negInts)), rep(1, nrow(posInts)))
  ret2 <- cbind(ret,HostTraits[(ret[,1]),], ParasiteTraits[(ret[,2]),])
  return(list(intMat = intmat, Hosts = HostTraits, Parasites = ParasiteTraits, pointsObject = ret2))
}


```







```{r eval = TRUE, echo = F, message = FALSE, warning = FALSE}
expandCol <- function(dat, colName = NULL){
  namez <- na.omit(unique(dat))
  ret <- sapply(namez, function(x){as.integer(grepl(x, dat))}) 
  colnames(ret) <- paste(colName, '.', namez, sep = '')
  return(as.data.frame(ret))
}

```



















```{r eval = TRUE, echo = TRUE, message = FALSE, warning = FALSE}


datBats <- filter(dat, host_order == 'Chiroptera')

host$host_species <- paste(host$MSW05_Genus, host$MSW05_Species, sep = ' ')

notHosts <- host$host_species[-which(host$host_species %in% unique(datBats$host_species))]
## interaction matrix
intMatFull <- as.data.frame.matrix(table(c(datBats$host_species, notHosts), c(datBats$virus_genus, rep('dummy', length(notHosts)))))

intMatFull$dummy <- NULL


### created features
# numParasites <- rowSums(intMatFull > 0)
numHosts <- colSums(intMatFull > 0)



h <- data.frame(host_species = rownames(intMatFull))
h <- left_join(host, h, by = 'host_species')
h <- left_join(h, cites, by = c('host_species' = 'name'))
h$X <- NULL
h$Pan <- NULL






### forming data.frame objects
host2 <- data.frame(
  expandCol(h$MSW05_Genus, 'hGenus'), 
  expandCol(h$ForStrat.Value, 'hForStratValue'),
  expandCol(h$IUCN , 'hIUCN '),  
  h[,c(6:60, 62:68)] 
  #	numParasites
)

host2$X12.2_Terrestriality <- NULL
host2$Diet.Vunk<- NULL
host2$Diet.Scav <- NULL
host2$Activity.Nocturnal<- NULL 
host2$X12.1_HabitatBreadth<- NULL
host2$Diet.Vfish<- NULL
host2$Diet.Nect <- NULL
host2$Diet.Seed<- NULL
host2$Diet.PlantO<- NULL
host2$urbc_1970AD<- NULL
host2$urbc_1980AD<- NULL
host2$urbc_1990AD<- NULL
host2$urbc_2000AD<- NULL
host2$urbc_2005AD <- NULL
host2$hFamily.DELPHINIDAE <- NULL
host2$ForStrat.Value <- NULL
host2$IUCN <- NULL

host2$BodyMass.value <- NULL
host2$PET_mean_mm <- NULL
host2$HuPopDen_5p_n.km2 <- NULL
host2$HuPop_Den_min_n.km2 <- NULL
host2$host_species <- NULL



virus2 <- datBats %>%
  group_by(virus_genus) %>%
  summarise(nHostSp = length(unique(host_species)),
            nHostFam = length(unique(host_family)),
            nHostOrder = length(unique(host_order)))

virus2$virus_genus <- NULL

## Points object
hvDat <- makePoints(intMatFull, host2, virus2)

```



```{r}
writeLinks <- function(dat, model, fileName = 'DallasPredictions.csv'){
  linksDF <- data.frame(host = rownames(dat[[1]])[model[[1]]],
                        virus = colnames(dat[[1]])[model[[2]]], 
                        presence = model[[3]]$presence,
                        suitability = model[[3]]$suitability)
  linksDF <- linksDF[order(linksDF$suitability, decreasing = TRUE),]
  ret <- linksDF[which(linksDF$virus  ==  'Betacoronavirus'), ]
  write.csv(ret, file = fileName)
}
```











```{r eval = TRUE, echo = TRUE, message = FALSE, warning = FALSE}

hvDati <- getImpute(hvDat, m = 5, useMeans = TRUE)

```





3.       Filter (host_order = CHIROPTERA) > PNP with trait approach (no latent) >  include citation counts > Return a csv of bat species and probability of interaction with virus_genus = ‘Betacoronavirus’ as outcome variable > DallasBatsCitations.csv


```{r eval = TRUE, echo = TRUE, message = FALSE, warning = FALSE}

hvOut1 <- getReal(hvDati[[4]], cv = FALSE)
writeLinks(dat = hvDati, model = hvOut1, fileName = 'DallasBatsCitations.csv')

```





4.       Filter (host_order = CHIROPTERA) > PNP with trait approach (no latent) >  don’t include citation counts > Return a csv of bat species and probability of interaction with virus_genus = ‘Betacoronavirus’ as outcome variable > DallasBatsUncorrected.csv


```{r eval = TRUE, echo = TRUE, message = FALSE, warning = FALSE}
hvDati[[2]]$cites <- NULL
hvDati[[4]]$cites <- NULL
hvOut2 <- getReal(hvDati[[4]], cv = FALSE)
writeLinks(dat = hvDati, model = hvOut2, fileName = 'DallasBatsUncorrected.csv')
```












































### The approach we took in the Open Science article (just use dissimilarity values)


```{r}
#' Format the data from a list of interaction matrices to a data.frame 
#' that can be input to the plug and play model.
#'
#'
#' @param intMatList list of interaction matrices
#' @param year year of prediction
#' @param lagYears number of years which to train the model on
#' @param distFun how to calculate dissimilarity values

formatPPobject <- function(intMat,
                           distFun = 'bray', includeCitations = FALSE){
  require(plyr)
  require(dplyr)
  locSim <- rowMeans(as.matrix(vegan::vegdist(intMat)))	
  pathSim <- rowMeans(as.matrix(vegan::vegdist(t(intMat))))
  simMat <- locSim %*% t(pathSim)	
  locDF <- data.frame(host = rownames(intMat), locSim)
  
  if(includeCitations){
    locDF <- left_join(locDF, cites,by = c('host' = 'name'))
    locDF$X <- NULL
  }
  
  pathDF <- data.frame(pathogen = colnames(intMat), pathSim)
  df <- expand.grid(rownames(intMat), colnames(intMat))
  colnames(df) <- c('host', 'pathogen')
  df$presence <- as.vector(intMat > 0)+0 
  df$similarity <- as.vector(simMat)
  df <- left_join(df, pathDF, by = 'pathogen')
  df <- left_join(df, locDF, by = 'host')
  return(df)
}

```










1.       Entire dataset > PNP with latent approach (no traits) > include citation counts > Return a csv of mammal species and probability of interaction with virus_genus = ‘Betacoronavirus’ as outcome variable > DallasMammalsCitations.csv

```{r}

intMatFull <- as.data.frame.matrix(table(dat$host_species, dat$virus_genus))
test <- formatPPobject(intMatFull, includeCitations = TRUE)
out <- getReal(test, cv = FALSE)

mamLinks <- cbind(test, suitability = out$predictions[,'suitability'])
mLinksCites <- mamLinks[which(mamLinks$pathogen  ==  'Betacoronavirus'), c('host','pathogen', 'presence', 'suitability')]
colnames(mLinksCites)[2] <- 'virus'
write.csv(mLinksCites, file = 'DallasMammalsCitations.csv')

```












2.       Entire dataset > PNP with latent approach (no traits) > don’t include citation counts > Return a csv of mammal species and probability of interaction with virus_genus = ‘Betacoronavirus’ as outcome variable > DallasMammalsUncorrected.csv

```{r}

test2 <- formatPPobject(intMatFull, includeCitations = FALSE)
out2 <- getReal(test2, cv = FALSE)

mamLinksNoCites <- cbind(test, suitability = out2$predictions[,'suitability'])
mLinksNoCites <- mamLinksNoCites[which(mamLinksNoCites$pathogen  ==  'Betacoronavirus'), c('host','pathogen', 'presence', 'suitability')]
colnames(mLinksNoCites)[2] <- 'virus'
write.csv(mLinksNoCites, file = 'DallasMammalsUncorrected.csv')

detach(package:plyr)

```
